---
title: "CatVar_FactAssig_Castelan"
author: "Castelan Moreno, Emily"
date: "2022-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stringr)
library(tidyverse)
library(forcats)
library(ggplot2)
```

# Please complete all the tasks listed below. After completing the tasks download the .Rmd file and upload in the peer review item for grading.

# Additionally please write text between the code chunks explaining what each code chunk is about.

# Refer the linked online textbook in case of any issues.


1.From the "forcats" library load gss_cat data. Explore the distribution of 
rincome (reported income). What makes the default bar chart hard to understand? 
How could you improve the plot?
```{r, question1}
## insepct the database (commented out)
## view(gss_cat)
## view the default bar chart 
## using rincome as the independent variable
## the labels are not consistent 
ggplot(gss_cat, aes(rincome))+
  geom_bar()
## we can see that the default bar chart is hard to understand because 
## the labels for income brackets are layered over one another 
## the values are so different that the values count >2000 are hard to gauge
## TO IMPROVE THE PLOT: 
##    1) move the labels to be vertical or angled to be legible 
##    2) rename the columns to be more consistent 
##    3) why is there a NA and a separate No answer? 
##       remove the NA because it throws off the other data 
##    4) rename the x and y axis 
##    5) reorder the columns from don't know, to least, to most 
rep_income_plot <- gss_cat %>%
  filter(!rincome %in% c("Not applicable")) %>%
    mutate(rincome = fct_recode(rincome, 
                              "Less than $1000" = "Lt $1000")) %>%
  ggplot(aes(rincome))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete("Reported Income in $") + 
  scale_y_continuous("Number of Respondents") 

rep_income_plot



```


2. What is the most common religion? What is the most common partyid?


```{r, question2}
## part 1
pop_relig <- gss_cat %>%
  group_by(relig) %>%
  summarize(relig_count = n()) %>%
  arrange(desc(relig_count))

pop_relig
## 1) group_by to focus on one column for calculations
## 2) use summarize and n() to count the observations based on the group 
## PROTESTANT is the most common religion

## part 2
pop_party <- gss_cat %>% 
  group_by(partyid) %>%
  summarize(party_count = n()) %>%
  arrange(desc(party_count)) %>%
  head(1)

pop_party
## 1) group_by to focus on the partyid column 
## 2) summarize, just like in the top code 
## 3) arrange the new column in descending order 
## 4) use head() to only print out the most common partyid 
## Independent is the most popular
  
```
3.Which relig does denom (denomination) apply to? How can you find out with a
table? How can you find out with a visualisation?

```{r, question3}
## first we can see denom is a column 
## so let's view that column using levels()
levels(gss_cat$denom)
## I don't know anything about religious denominations so I had to google this 
## but anyways apparently these are denom is for protestant 
## we can prove this using this table:
pop_denom <- gss_cat %>%
  group_by(denom, relig) %>%
  summarize(denom_count = n()) %>%
  arrange(desc(denom_count)) %>%
  filter(denom != "Not applicable") %>%
  filter(denom != "No denomination") %>%
  filter(denom != "No answer") %>%
  filter(denom != "Don't know")
pop_denom
## 1) first group by denom and religion
## 2) use count just to put some order to it 
## 3) rearrange in descending order 
## 4) filter out the religons where denom is not applicable 
## 5) inspect the table we can see the religion is protestant 

## now let's visualize this 

denom_plot <- gss_cat %>%
  group_by(denom, relig) %>%
  summarise(denom_count = n())
denom_plot %>%
  ggplot(aes(x = relig, y = denom, size = denom_count)) + 
  geom_point() +
  theme(axis.text.x = element_text(angle = 90)) 

denom_plot

## 1) use the same group+by and summarize code block 
## 2) plot those results 
## 3) I tried color and size to show the counted denom_count, but size shows 
##    the results the best 
## 4) Showing once again that denomination is used for the Protestant religon

```

4.How have the proportions of people identifying as Democrat, Republican,
and Independent changed over time? Plot a suitable chart.
```{r, question4}
## first organize the data 
party_history <- gss_cat %>%
  group_by(year, partyid) %>%
  summarise(party_count = n()) 

party_history %>%
  ggplot(aes(x = year, y = party_count, color = partyid)) + 
  geom_point() +
  geom_line()
  
party_history

## HOWEVER the question says Democrat, Republican and Independent
## so let's collapse the party affilication observations into those categories
## ALSO it asked for the proportion so we need to turn it into percentage 

party_simple <- gss_cat %>%
 mutate(
    partyid2 =
      fct_collapse(partyid,
        other = c("No answer", "Don't know", "Other party"),
        republican = c("Strong republican", "Not str republican"),
        independent = c("Ind,near rep", "Independent", "Ind,near dem"),
        democrat = c("Not str democrat", "Strong democrat")
      )
  ) 
## 1) custom define the groups using fct_collapse()
party_simple <- party_simple %>% 
  group_by(year, partyid2) %>%
  summarise(party_count2 = n()) %>%
  mutate(party_proportion = party_count2/sum(party_count2))
## 2) group by year and party id and get a cound using summarize
## 3) mutate a new column that calcuates the proportion 
##    based on the party_count, party_count/total(party_count)

party_simple %>% 
  ggplot(aes(x = year, y = party_proportion, color = partyid2))+
  geom_point()+
  geom_line()
## 4) plot the results 
## 5) use geom_line to connect the points across the years 

## overtime, 
## independent has increased a little 
## democrat has stayed relatively level 
## republican has decreased
## and other has seen the least amount of change
```
5.Collapse "rincome" into smaller set of categories?
```{r, question5}
## first inspect rincome levels 
levels(gss_cat$rincome) 
## rename the levels manually
rincome_simple <- gss_cat %>%
  mutate(
    rincome2 = 
      fct_collapse(rincome, 
                   'Unknown' = c("No answer", "Don't know", "Refused", 
                               "Not applicable"), 
                   'Less than $10000' = c("Lt $1000", "$1000 to 2999",
                                         "$3000 to 3999", "$4000 to 4999", 
                                         "$5000 to 5999",  "$6000 to 6999",
                                         "$7000 to 7999", "$8000 to 9999"), 
                  '$1000 - $19999' = c("$10000 - 14999", "$15000 - 19999"),
                  '$20000 to 24999' = c("$20000 - 24999"),
                  'more than $25000' = c("$25000 or more")
      )
  )
## plot the new rincome using a bar graph

  rincome_simple %>%
    ggplot(aes(x = rincome2)) +
    geom_bar()



```

