---
title: "StringLab_Castelan"
author: "Castelan Moreno, Emily"
date: "2022-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(tidyverse)
library(htmlwidgets)

```

##### Assignment Instructions

Complete all questions below. After completing the assignment, knit your
document, and download both your .Rmd and knitted output. Upload your files 
for peer review. 

For each response, include comments detailing your response and what each
line does. Ensure you test your functions with sufficient test cases to
identify and correct any potential bugs.

##### Question 1.

Use str_c to put `(` before the area codes followed by `)` and a space followed
by the phone number. 
```{r, question1}
### Answer should be of the form "(703) 5551212" "(863) 1234567" 
### "(404) 7891234" "(202) 4799747"

area_codes <- c(703, 863, 404, 202)
phone_nums <- c(5551212, 1234567, 7891234, 4799747)

str_c("(",area_codes, ") ", phone_nums)

## 1) concatenate each 
## 2) str_c to join the two vector elements
##    specify and added "( )" around the first vector element
```

##### Question 2.

Create a function that receives a single word as an input. Use str_length() and
str_sub() to extract the middle character from the string. What will you do if
the string has an even number of characters? Test your function on the strings
“hamburger” and “hotdog”

```{r, question2}
## test cases
x <- "hamburger"
y <- "hotdog" 
s <- "lo"
name <- "jesse"
w <- c("hot, hot, hot")


## create function that finds the median point of a single string 
## ** it will only accept 1 argument 
median_string <- function(x){
  if(
    str_detect(x, ",") == TRUE){warning("only single word argument accepted")}
  else{
     Leng <- str_length(x)
     if(Leng <= 2){str_sub(x, 1, 1)}
     else{
     Med <- ceiling(Leng/2)
    str_sub(x, Med, Med)}
  }
}
## 1) it checks if the argument is a single word 
## 2) if it is, it will take the length 
## 3) divide the length by 2 to find median 
## 4) if even number, takes ceiling (takes smallest integer)
## 5) if length is equal to or less than 2, returns first letter
## 6) str_sub() used to remove the start and end of the word to find median chr

## test each sample case
median_string(x)
median_string(y)
median_string(s)
median_string(w)
## the last case is designed to spit out an error 
```
##### Question 3. 

How would you match the sequence "’\? Note this is a double quote, single quote,
backslash and question mark. Build it up one piece at a time. 
Use it to identify that sequence contained in s2 below.
```{r, question3}
## I realized after the fact this is not what the question is asking for: 
s <- " \"'\\? "
s2 <- str_c("some stuff", s ,"more!")


#writeLines(s)
#writeLines(s2)
writeLines(s)
writeLines(s2)

## 1) use \ to escape the "" 
## 2) use a second \ to differentiate beween the printed \ 
## 3) use a " to end the function 

## THE CORRECT ANSWER: 
str_view("\"'\\?", "\"'\\\\.{0,1}", match = TRUE)
str_view(s2, "\"'\\\\.{0,1}", match = TRUE)


```
##### Question 4. 

Using the words provided in stringr::words, create regular expressions 
that find all words that:
```{r, question4}
# End with "ing" or "ise"
writeLines("words ending in \"ing\" or \"ise\":\ ")
## just to clean up the print 
str_subset(stringr::words, "(ing|ise)$")
## we want words that: 
##      end with ing
##      OR |
##      end with ise 
##      we use the $ outside of either case presented


# Do not follow the rule "i before e except after c"
writeLines("words that break rule: \"i before e except after c:\ ")
str_subset(stringr::words, "ei|cie")
x <- as_tibble(str_subset(stringr::words, "ei|cie"))
count(x)
## this means finds words with: 
##        words with e before i 
##     OR words with c-i-e
## count the number of words using count
## there are 6 words that do not follow the rule


# Begin with at least two vowels and end with at least two consonants
writeLines("words that: begin with 2 vowels and end with 2 consonants: ")
str_subset(stringr::words, "^[aeiou]{2}.*[^aeiou]{2}$")
## we want words that: 
##      start^ with [aeiou] at least {2} 
##      and .* 
##      end with NOT[^aeiou]$ ($ is the end specifier), at least {2}
##        NOTE: we use {2} before the $ otherwise it returns words like "out"


# Contain a repeated pair of letters (e.g. "church" contains "ch" twice)
writeLines("words contain a repeat pair of letters: ")
str_subset(stringr::words, "([A-Za-z][A-Za-z]).*\\1")
## we want words that: 
##      have two letters which are A-Z and a-z 
##          +using case sensitive 
##      and .* 
##      match the pattern, using \\1 backreference, 
##      aka repeats the first pair 


# Contain one letter other than e that is repeated in at least three places 
# (e.g. “appropriate” contains three “p”s.)
writeLines("words that contain a letter repeated 3x anywhere, not e:")
str_subset(stringr::words, "([a-df-z]).*\\1.*\\1")
## we want words that use the same letter a-d + f-z 3 times anywhere
##            skip over e!
##      like before set the possible character to any letter a-z
##      use \\1 again to back reference the previous letter twice 
##      use .* to specify AND 
```

##### Question 5. 

Using the sentences provided in stringr::sentences, find all words that come
after a “number” like “one”, “two”, … “twelve”. Pull out both the number and 
the word.
```{r, question5}
## first attempt (commented out)
## written_nums <- c("one", "two","three", "four", "five", "six", "seven", 
##              "eight", "nine", "ten")
## num_match <- str_c(written_nums, collapse = "|")
## has_num <- str_subset(stringr::sentences, num_match)
## matches <- str_extract(has_num, num_match)
## head(matches)
## more <- sentences[str_count(sentences, num_match) > 1]
## str_view(more, num_match)
## THE PROBLEM ^^^ 
##          this returns all numbers within other words 
##          we want all numbers (not within another word)
##          the word that follows the number 

## second attempt: 

num_as_word <- "\\b(zero|one|two|three|four|five|six|seven|eight|nine|ten
                    eleven|twelve|twenty|thrity|fourty|fifty|sixty
                    seventy|eighty|ninety|hundred) +(\\w+)"
sentences[str_detect(sentences, num_as_word)] %>%
  str_extract(num_as_word)

## 1) specify the written numbers using \\b which matches word boundaries 
##      the boundaries are the numbers written out 
##      (\\w+) matches any "word" character following the number 
```

##### Question 6. 

Using the sentences provided in stringr::sentences, view all sentences that
contain the word “good” or the word “bad”. Get the sentence numbers where those
words occur. Use str_replace_all() to replace the word “bad” with the word 
“horrible” and the word “good” with the word “great”. Look at the sentence 
numbers you found before to verify the words were replaced correctly.
```{r, question6}
## FIRST idea: (commented out)
## bad_good <- "(\\w+) +\\b(good|bad) +(\\w+)"
## sentences[str_detect(sentences, bad_good)] %>%
##   str_extract_all(bad_good)
## THE PROBLEM ^^^
##      This does not return the whole sentences 
##      This does not keep the sentence number 
##      Remove the last line and it keeps the sentence intact
##      BUT it still changes the sentences numbering 

bad_good <- "(\\w+) +\\b(good|bad) +(\\w+)"
bad_good_sent <- sentences[str_detect(sentences, bad_good)]
## 1) detect for the words "bad" and "good" in sentences 
bad_good_sent
bad_good_sent <- str_c(bad_good_sent, collapse = "|")
bad_good_sent
## 2) collapse it into a single vector with | as the separator

sentence <- as_tibble(sentences)
## 3) convert  into a tibble so sentences are numbered 

which(grepl(bad_good_sent, sentence$value))
## 4) this returns the row number of the sentences that have "bad" or "good"

## 5) use str_replace_all() 
##    change bad to horrible 
##    change good to great
```

