---
title: "Tidying_DataLab_Coursera_Castelan"
author: "Castelan Moreno, Emily"
date: "2022-11-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```
# Please complete all the tasks listed below. After completing the tasks download the .Rmd file and upload in the peer review item for grading.

# Additionally please write text between the code chunks explaining what each code chunk is about.

# Refer the linked online textbook in case of any issues.
##### Question 1. 

The built in billboard dataset is not tidy. Describe why it is not tidy and then 
tidy the dataset.
```{r, question1_response}
## Billboard is not tidy because the weeks columns are should be observation 
## values, and the number of weeks does not correspond to a value from the date

## First gather up all the week entries into a row for each week for each song 
## (where there is an entry)
billboard 
##print billboard to compare

billboard_tidy <- billboard %>% 
  pivot_longer(
    cols = 'wk1':'wk76',
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )


# Then, convert the week variable to a number and figure out the date 
## corresponding to each week on the chart
billboard_tidy <- billboard_tidy %>%
    mutate(
    week = as.integer(gsub("wk","",week)),
    date = as.Date(date.entered)+7*(week-1),
    date.entered = NULL
  )

billboard_tidy
## I dropped the date entered since the date entries is the same as the 
## first week it seems unnecessary to keep both columns... 

# Sort the data by artist, track and week. Here are what your first entries 
## should be (formatting can be different):
#>  A tibble: 5,307 x 5
#   artist  track                   date.entered  week  rank   date
 #    <chr>   <chr>                   <date>       <int> <dbl>   <date>
 #  1 2 Pac   Baby Don't Cry (Keep... 2000-02-26       1    87 2000-02-26
 #  2 2 Pac   Baby Don't Cry (Keep... 2000-02-26       2    82 2000-03-04
 #  3 2 Pac   Baby Don't Cry (Keep... 2000-02-26       3    72 2000-03-11
 #  4 2 Pac   Baby Don't Cry (Keep... 2000-02-26       4    77 2000-03-18
 #  5 2 Pac   Baby Don't Cry (Keep... 2000-02-26       5    87 2000-03-25
 #  6 2 Pac   Baby Don't Cry (Keep... 2000-02-26       6    94 2000-04-01
 #  7 2 Pac   Baby Don't Cry (Keep... 2000-02-26       7    99 2000-04-08
 #  8 2Ge+her The Hardest Part Of ... 2000-09-02       1    91 2000-09-02
 #  9 2Ge+her The Hardest Part Of ... 2000-09-02       2    87 2000-09-09
 # 10 2Ge+her The Hardest Part Of ... 2000-09-02       3    92 2000-09-16
 # … with 5,297 more rows
```

##### Question 2. 

Tidy the "fish_encounters" dataset of fish spotting by monitoring stations.
Make the NA into 0 using the option "values_fill = list(seen = 0)"

```{r, question2_response}
## load the dataset to compare
fish_encounters


fish_tidy <- fish_encounters %>% 
  pivot_wider(
    names_from = station,
    values_from = seen, 
    values_fill = list(seen = 0)
  )
  
fish_tidy
## we made it so there is only one row per fish number 
## and filled in NA's with 0 
```
##### Question 3. 

Import the flowers1 dataset. Tidy and pivot the data. Hint: use "read_csv2()" 
to read in the dataset 

```{r, question3_response}
## download the file from the URL, 
## save it in a folder
## read it in using csv2
fileUrl <- "https://raw.githubusercontent.com/JaneWall/STAT412_612_online/master/Week%208%20tidyr/Data/flowers1.csv?accessType=DOWNLOAD"

if(!file.exists("Data")){dir.create("Data")}

download.file(fileUrl, destfile ="Data/flowers1.csv")
(dateDownloaded <-date())

flowers1 <- read_csv2(fileUrl)
flowers1

## tidy the dataset 
flowers1_tidy <- flowers1 %>%
pivot_wider(
  names_from = Variable,
  values_from = Value
)
flowers1_tidy
```
##### Question 4.

Import the flowers2 dataset. Tidy the dataset by turning the one column into 3
separate columns
```{r, question4_response}
## download the dataset 
## save it in a folder 
## read it in 
fileUrl2 <- "https://raw.githubusercontent.com/JaneWall/STAT412_612_online/master/Week%208%20tidyr/Data/flowers2.csv?accessType=DOWNLOAD"

if(!file.exists("Data")){dir.create("Data")}

download.file(fileUrl2, destfile = "Data/flowers2.csv")
(dateDownloaded <- date())

flowers2 <- read_csv2(fileUrl2)

flowers2

## now tidy the dataset
## there are two values in the same column we will use separate()
## because there is a decimal point and a / in the column, we must 
## specify the separater 
flowers2_tidy <- flowers2 %>%
separate('Flowers/Intensity', into = c("Flowers", "Cases"), sep = '/')
      
flowers2_tidy
```

##### Question 5.

In the following dataset, turn the implicit missing values to explicit 

```{r, question5_response}
output <- tibble(
      treatment   = c("a", "b", "a", "c", "b"),
      gender   = factor(c("M", "F", "F", "M", "M"), levels = c("M", "F", "O")),
      return = c(1.5, 0.75,  0.5,  1.8,  NA)
    )
output

## there should be a treatment for each gender: 
## F-a, F-f, F-c
## M-a, M-b, M-c, 
## but instead it is missing values for F-c and M-b
## let's make the implicit values missing using pivot_wider: 

output %>% 
  pivot_wider(names_from = treatment, values_from = return)

## there are two NA values now in the cells we said were implicitly missing

## this question is a little unclear as to what "O" is but here is another 
## possible solution using "O" as a gender option 
## I assumed because there were no values for "O" that it could be ignored
## but if we want to return an NA for each "O" then.. 

output_two <- tibble(
      treatment   = c("a", "b", "a", "c", "b"),
      gender   = factor(c("M", "F", "F", "M", "M"), levels = c("M", "F", "O")),
      return = c(1.5, 0.75,  0.5,  1.8,  NA)
    )
output_two %>% complete(treatment, gender)


```
#####Question 6.

Import the weather dataset as weather. Use "pivot_longer()" to put the days all
in one column, then use "pivot_wider" to separate tmax and tmin into separate
columns. Print the summary of the final resulting dataset 
```{r, question6_response}

fileUrl <- "https://raw.githubusercontent.com/JaneWall/data_STAT412612/master/weather.csv?accessType=DOWNLOAD"

## downlaod the file 

 if(!file.exists("Data")){dir.create("Data")}
## create a folder for the file if one does not exist 

download.file(fileUrl, destfile = "Data/weather.csv")

(dateDownloaded <- date())
## put a date on the download 

weather <- read_csv(fileUrl)
weather
## read in as csv because it is comma delimited

## now to tidy the data
weather_fixed <- weather %>%
pivot_longer(
  cols = d1:d31,
  names_to = "day", 
  values_to = "temp", 
  values_drop_na = TRUE)
## first we see that the variables d1:d31 should be observations not columns 
##  use pivot_()
##  specify the columns to be changes into observations 
##  set the names_to for the new column name 
##  set the column where we will place the values that were inside d1:d31
## the column to take variable names from is "element" for names_from
## the column to take values from is 
weather_fixed <- weather_fixed %>% 
  pivot_wider(
  names_from = element, 
  values_from = temp)

weather_fixed
```

###### Question 7. 

Load the built in "anscombe" data frame and use "pivot_longer()" to separate
all the x and y columns and categorize them into 4 sets
```{r, question7_response}
## load in the dataset to compare 
anscombe <- as_tibble(anscombe)

anscombe_tidy <- anscombe %>% 
 pivot_longer(cols = everything(),
             names_to = c(".value", "set"),
             names_pattern = "(.)(.)")

anscombe_tidy

## first use pivot longer to consolidate the columns 
## specify the columns to send the values to with names_to 
## use names_pattern() to specify how column names should be split 
## note using the ".value" overrides the values_to 

```

