---
title: "Data_Analysis_Lab_Castelan"
author: "Castelan Moreno, Emily"
date: "2022-10-31"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)

```

##### Assignment Instructions

Complete all questions below. After completing the assignment, knit your 
document, and download both your .Rmd and knitted output. Upload your files for
peer review. 

For each response, include comments detailing your response and what each 
line does. 

---

##### Question 1. 

Using the nycflights13 dataset, find all flights that departed in July, August,
or September using the helper function between(). 


```{r, question1_response}
## use a pipe to input the dataset, set an object that wills save our result
july_to_sept <- flights %>% 
  filter(between(month, 7, 9))
## use a filter() followed by between()
## define what is being filtered, the left lower end and, right upper end 
## returns a tibble 86,326 x 19 

total_with_canceled <- count(july_to_sept)
## use a count to return the number of flights from july to sept. 
## to account for flights that were canceled, 

not_canceled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
## create an object that store the result of filtering out flights with NA 
## values for the arr_delay or dep_delay

## now use this not_canceled to compare against our july to sept.filter 
## store it in july_to_sept2
july_to_sept2 <- not_canceled %>%
  filter(between(month, 7, 9))

## now compare the two different values, one accounting for canceled flights, 
## and the other not 

total_wo_canceled <- count(july_to_sept2)
print(total_wo_canceled)

## there are about 2,000 less flights from july to sept. when we account for 
## the cancelled flights 
  
## divide the two to get the % 

perct_same <- (total_wo_canceled / total_with_canceled)*100
print(perct_same)

```

#### Question 2. 

Using the nycflights13 dataset sort flights to find the 10 flights that flew the
furthest. Put them in order of fastest to slowest.

```{r, question2_response}
## 1st, sort flights by air_time and store the result 
furthest_bytime <- flights %>%
  arrange(desc(distance)) %>% 
  ## pipe that resulting table into a select() for the important variables
     select(distance, air_time, year, month, day, flight) %>%
  ## pipe that into a mutate() to create a new variable 
  ## use distance = rate x time to solve for the rate and store it 
  ##      therefore rate = distance / time 
        mutate(rate = distance / air_time) %>% 
  ## arrange the table in descending order by the new column rate 
  arrange(desc(rate))
## print only the top 10 using head()
head(furthest_bytime, n = 10)

```

#### Question 3. 

Using the nycflights13 dataset, calculate a new variable called "hr_delay" and 
arrange the flights dataset in order of the arrival delays in hours (longest
delays at the top). Put the new variable you created just before the departure 
time.Hint: use the experimental argument .before. 

```{r, question3_response}
## first pipe the flights dataset into the 
flights_hr_delay <- flights %>%
  ## make a new variable called hr_delay which is the arr_delay in hours
  mutate(hr_delay = arr_delay/60) %>% 
  ## arrange that variable in descending order 
  arrange(desc(hr_delay)) %>% 
  ## reclocate the placement of hr_delay to before dep_delay
  relocate(hr_delay, .before = dep_delay)
  ## just to make things more clear round the values in hr_delay to 2 digits 
  ## rewrite over hr_delay, specific the column in the dataset with $
  flights_hr_delay$hr_delay <- round(flights_hr_delay$hr_delay, digit = 2)
  
  ## return the results 
  print(flights_hr_delay)

```

##### Question 4.

Using the nycflights13 dataset, find the most popular destinations 
(those with more than 2000 flights) and show the destination, the date info, 
the carrier. Then show just the number of flights for each popular destination.

```{r, question4_response}
flights_dest <- flights %>% 
  ## group the flights by destination 
  group_by(dest) %>% 
  ## use filter() to filter for groups of destinations greater than 2000
  filter(n() > 2000) %>%
  ## show only the destination, the date info, the carrier 
  select(year:day, dest, carrier)

## at this point typing in "flights_dest"
  ## this returns a tibble 302,969 x 5
  ## groups: dest [46]
## BUT WE WANT: 
  ## SHOW JUST the number of flights for each popular destination 
flights_dest %>%
  summarize(count = n(), 
            .groups = "drop_last") %>% 
  arrange(desc(count))

## We used summarize to collapse the data frame based on the number 
##      of flights for each destination 
## set count = n() to count by destination group 
## arrange in descending order to check that all counts are greater that 
## 2000 flights 
```

#### Question 5. 

Using the nycflights13 dataset, find the flight information (flight number, 
origin, destination, carrier, number of flights in the year, and percent late) 
for the flight numbers with the highest percentage of arrival delays. Only
include the flight numbers that have over 100 flights in the year.

```{r, question5_response}
flights_per_year <- not_canceled %>%
  select(flight, arr_delay, origin, dest, carrier) %>%
  group_by(flight) %>% 
  summarize(count = n(), 
            across()) %>% 
  filter(count > 100) %>% 
  ## find number of flights in year by flight number 
  ## filter for only flights that have greater than 
  ## 100 flights in a year
  select(origin, dest, carrier, count, arr_delay, 
         flight) %>% 
  ## select only the information needed
  
  group_by(flight) %>%
  mutate(
    delay_total = sum(arr_delay), 
    prop_delay = arr_delay / delay_total *100) %>% 
  filter(delay_total != 0) %>%
  filter(arr_delay > 0) %>%
  arrange(desc(prop_delay), flight) 

    
## calculate the perct. arrival delay 
## filter out cases where the flight was not delayed 
## filter out cases that equal zero for delay_total
## arrange by greatest proportion for arrival delay by flight


flights_per_year


```









