---
title: "PracticeProblemsSet1_Castelan"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Review Sections 2.1-2.5 from Hands-On Programming in R 
1. Summary of what I learned: 
  colon operator : , returns every integer between two integers 
  ls() , function which prnts object names already used 
  R uses element-wise execution by default 
  If R is provided two vectors of unequal lengths, it will repeat the shorter 
  until it is the same length of the longer vector, this is called 
  vector recycling 
  Matrix operators: 
    %*% inner multiplication, dot product 
    %o% outter multiplication 
  look up function's arguments with, args(name_of_function)
  sample with replacement:
    sample(die, size = 5, replace = TRUE)
    created independent random samples, otherwise default is without 
    replacement for a sample 
  arguments can be set to defaults as opposed to specified for each time 
  the function is called 
    functions can have as many arguments as needed/desired 
2. Add a second chunk to your document that contains the roll2 functions.
   Knit your document to be sure it does not have errors. 
```{r, Roll2Function}
# function which randomly generates the sum or "dice" which is in this case 
# "bones" the argument which is specified as values 1 through 20 
# by randomly selecting two values without replacement for each selection 
# from the sample (set of potential outputs)
roll2 <- function(bones = 1:20){
  dice <- sample(bones, size = 2, replace = FALSE)
  sum(dice)
}
```
   
3. Think about what sort of inputs should be allowed for roll2 and write in 
text section what the input must look like. Test roll2 both without and input 
to be sure the default is used as well as with several different inputs. Be 
sure your document shows the results of your tests when it is knitted. 

```{r, Roll2Tests, error = TRUE}
# the function above roll2, requires bones to: 
# 1) have at least two values because the size = 2 
# 2) cannot be a character type, must be numeric type 
# 3) sample cannot take a sample larger than the population when replace = FALSE
#    therefore bones cannot equal Inf 

# Demonstrating different tests of inputs for argument bones
# Test 1 without specifiying a default 

roll2 <- function(bones = 1:20){
  dice <- sample(bones, size = 2, replace = FALSE)
  sum(dice)
}
roll2()

# Test 2, bones as a character argument 
roll2(bones = c("dog", "bunny", "cat"))

# Test 3, bones as infinity 
roll2(bones = Inf)

# Test 4, bones as a value less than size 2
roll2(bones = 1)

```

# Review Sections 3.1-3.4 from Hands-On Programming in R 
## and complete the following exercises: 

1. In a new code chunk in your Rmd file, produce a histogram of 50,000 rolls 
of three 8-sides fair dice

```{r, FairDiceFunction}
# set up a function which takes the type of dice as an argument, but we set 
# the default to an 8 sided die, then in the sample we specify the parameters

RollDice <- function(Die8Sided = 1:8){
  dice8Sided <- sample(Die8Sided, size = 3, replace = TRUE)
  sum(dice8Sided)
}

# Use replicate to execute the function 50,000 times and store it in a vector 

NumberofRolls <- replicate(50000, RollDice())

# use qplot to create a histogram wih the NumberofRolls
qplot(NumberofRolls, binwidth = 1)


```
2. Then do the same thing where the dice are loaded so that the number 7
has a higher probability of being rolled than the other numbers, assume all the
numbers have a 1/10 probability of being rolled 

```{r, RiggedDice}
# first set up the function which takes the type of dice as an argument 

RiggedDice <- function (WeightedDice = 1:8){
  DiceWeighted <- sample(WeightedDice, size = 3, replace = TRUE, 
    prob = c(1/10, 1/10, 1/10, 1/10, 1/10, 1/10, 3/10, 1/10))
  sum(DiceWeighted)
}

# We will plot this function as well to see if it is executing correctly
# use qplot to create a histogram wih the RiggedDice
# first replicate the results 10000x to plug into histogram

RiggedDiceRolls <- replicate(10000, RiggedDice())

qplot(RiggedDiceRolls, binwidth = 1)

```


# 1. Rewrite the rescale01() function such that -inf is mapped to 0 and Inf is mapped to 1
```{r, rescale01, echo = TRUE}

#first set up the data frame, using the tibble function which is just a type of
# data frame in the tibble package 
# input vectors a, b, c, and d 
# within each vector use rnorm, which generates random numbers using a normal 
#   bell curve distribution 
# r norm syntax ----> rnom(n, mean, sd)
#   n = the number of random variables to generate 
#   mean = if not specified defaults to 0 
#   sd = the standard deviation, defaults to 1

df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

# Next, write the function to normalize the data 
# the only input is x, which will correspond to each vector a-d
# we can set range as a function "rng" which returns the range of x
#   we set na.rm = TRUE, meaning it will return NA from the calculation 
#   if an NA is input into the data frame
# lastly, inputing rng[1] means minimum and inputing rng[2] means maximum
#   which are in this case 0 and 1 

rescale01 <- function(x){
 rng <- range(x, na.rm = TRUE, finite = TRUE)
 (x-rng[1])/(rng[2]-rng[1])
ifelse (x == Inf, atan(x)*(2/pi),
  ifelse (x == -Inf, 1/(1+(x)), x))
}

# These modifications were made to the same function above after: 
# now we want to write the rescale01 function such the -Inf is mapped to 0 
# and Inf is mapped to 1 
# first set finite = TRUE 
# Then write two ifelse statements, because they accept vectors unlike 
# else if and if statements that require for loops to run vectors 
# THEN nest one ifelse, notice the syntax if left open on first ifelse
# finally, close both 
# note, using formulas that we know will return the desired values for scaling

# Test run the modified function 
y <- c(-Inf, -0.5, 3, 7, 100, Inf, .6, NA, NA)
rescale01(y)
# it works finally! 
```
# 2. Write both_na(), a function that takes two vectors of the same length
# and returns the number of positions that have an NA in both vectors 

```{r, both_na_function}
# both_na will begin by being a function with two inputs 
# in the body we will pass arguments that set NA to true and checks for NA 
# Then when the position of NA matches between the two vectors, it will count 
# the total number of NA's 

# test cases 
vector1 <- c(4, 7, NA, NA, NA, 8, 90, 8, NA)
vector2 <- c(NA, 9, 70, NA, 8, 900, 8, NA, NA)
counter_vector<- c(0)


both_na <- function(x, y){
  i <- which(is.na(x))
  j <- which(is.na(y))
  sum(ifelse(i == j, (counter_vector+1), counter_vector))
}

# recap of function 
# use which() to first determine where the NA values are (position) and store
# ifelse, to run both i & j vectors, count where the NA are in same position
#   if the vectors are the same, add 1 to the counter, if not keep the same
#   finally, run the sum

# run the test vectors
both_na(vector1, vector2)


```