---
title: "CapstoneCoursera_pt2"
author: "Castelan Moreno, Emily"
date: "2022-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(usmap)
library(ggplot2)
library(forcats)
library(zoo)
library(patchwork)
library(stringr)
library(ggrepel)
library(usmap)
library(cowplot)
library(gridExtra)
```
## THE DATASETS
```{r, import-nyt-data, include=FALSE}
# Import New York Times COVID-19 data
# Import Population Estimates from US Census Bureau 

us_counties_2020 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-2020.csv")
us_counties_2021 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-2021.csv")
us_counties_2022 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties-2022.csv")

us_population_estimates <- read_csv("fips_population_estimates.csv")
```

##### Part 2 - US State Comparison

While understanding the trends on a national level can be helpful in 
understanding how COVID-19 impacted the United States, it is important to 
remember that the virus arrived in the United States at different times. 
For the next part of your analysis, you will begin to look at COVID related 
deaths and cases at the state and county-levels.


##### Question 1

Your first task in Part 2 is to determine the top 10 states in terms of total
deaths and cases between March 15, 2020, and December 31, 2021.

Once you have both lists, briefly describe your methodology and your results.

```{r, part2_question1, echo=TRUE, include=TRUE}
# Determine the top 10 states in terms of total deaths and cases between 
# March 15, 2020, and December 31, 2021. To do this, transform your combined 
# COVID-19 data to summarize total deaths and cases by state up to 
# December 31, 2021.
## setting up our counties_combined again 
counties_combined <- rbind(us_counties_2020, us_counties_2021, us_counties_2022)

cutoff_date <- as.Date("2020-03-15")


counties_filtered <- counties_combined %>%
  filter(state != "Puerto Rico") %>%
  filter(date >= cutoff_date)

counties_filtered

## 1) view the us_population_estimates data set and determine the primary key 
## this code returns a 0 
primary_key <- us_population_estimates %>%
  count(fips, Year, Estimate) %>%
  filter(n>1)

## 2) set the second cutoff date as Dec 31 2021 
## **we want to set the date equal to the cutoff date because that is already 
## the total number of cases and deaths, whereas if we change use "<=" then 
## it will aggregate all the totals again and it will overcounted 
cutoff_date_dec21 <- as.Date('2021-12-31')
counties_filtered2 <-  counties_filtered %>%
  filter(date == cutoff_date_dec21)

## 3) use this new data set to find the total cases and deaths by state via 
## aggregate() 

state_cases <- 
  aggregate(counties_filtered2$cases
            ~ counties_filtered2$state,
            data = counties_filtered2,
           FUN = sum)


state_deaths <- 
  aggregate(counties_filtered2$deaths
            ~ counties_filtered2$state,
           data = counties_filtered2,
           FUN = sum)




## 4) Rename the columns to review the data without confusion and simplify next
## joining step 

state_deaths <- state_deaths %>%
  rename(
    state = 'counties_filtered2$state', 
    deaths = 'counties_filtered2$deaths'
  )

state_cases <- state_cases %>%
  rename(
    state = 'counties_filtered2$state',
    cases = 'counties_filtered2$cases'
  )

## 5) use an inner join to join the two totals by state 

by_state_total <- inner_join(
  state_deaths,
  state_cases,
  by = "state")

by_state_total<- by_state_total%>%
  filter(
    state != "American Samoa"
  )

## 6) I don't know why the output has the date, but let's add it back in just to 
## match it 
## 7) arrange the states in descending order, we can use two columns 
## as arguments

by_state_total <- by_state_total %>%
  mutate(
    date = cutoff_date_dec21
  ) %>%
  arrange(
    desc(deaths), cases
  )

## 8) take only the top 10 and print out the answer

top10_states <- by_state_total %>%
  head(n=10)

as_tibble(top10_states)

writeLines("10 states with highest COVID-19 deaths and cases, 
           in descending order")
print(top10_states$state)


```
-- Communicate your methodology, results, and interpretation here -- 
I initially coded this response wrong because I aggregated the total deaths and 
cases without filtering for only the cutoff date Dec. 31st 2021 which 
printed a result that was WAY OVER the correct values. This is because it was 
overcounting the data and it was apparent because a single state had the same 
number of cases for the whole US. Anyways, once I realized that I filtered the
data set for just the single date, which already has the total. Then from 
there I used aggregate() to collapse the rows by state for the cases variable 
then the deaths variable. I did this to verify at each step that it was done 
correctly then I joined the two results using an inner join. I added the 
date column back into my table, but I don't think it serves much of a purpose 
OTHER than to mark the date the data set is counting up to, which I guess could 
have helped me see my initial mistake. Next, I arranged the data set in 
descending order by deaths and cases as the two arguments and printed the top 10. 

##### Question 2

Determine the top 10 states in terms of deaths per 100,000 people and cases 
per 100,000 people between March 15, 2020, and December 31, 2021.


Once you have both lists, briefly describe your methodology and your results. 
Do you expect the lists to be different than the one produced in Question 1? 
Which method, total or per 100,000 people, is a better method for reporting 
the statistics? 

```{r, part2_question2, echo=TRUE, include=TRUE}
# Determine the top 10 states in terms of deaths and cases per 100,000 people
# between March 15, 2020, and December 31, 2021. You should first tidy and 
# transform the population estimates to include population totals by state. 
# Use your relational data verbs (e.g. full_join()) to join the population 
# estimates with the cases and death statistics using the state name as a key.
# Then, use case_when() and grepl() to add a population column to your table 
# that only includes the estimated population for the associated year. Finally,
# mutate your table to calculate deaths and cases per 100,000 people and 
# summarize by state. 

## 1) inspect and tidy the us_population_estimates
## as_tibble(us_population_estimates)
## 2) so let's group_by() and summarize by the state and year 
## otherwise we would get a total of the two years! and over count 
state_pop_est <- us_population_estimates %>%
  group_by(STNAME, Year) %>%
    summarise(pop_estimate = sum(Estimate))
state_pop_est <- state_pop_est %>%
  rename(
    state = 'STNAME'
  )

## 3) use the full_join() to join the table to our previous table 
## we will only need values for 2021 
top10_per100k <- full_join(
  state_pop_est,
  by_state_total,
  by = "state"
) %>%
  filter(Year == '2021')
  

## 4) divide the deaths and cases total by the corresponding Year for pop. 
## estimate for each state then multiple by 100,000


top10_per100k <- top10_per100k %>%
  mutate(
    deaths_per100k = deaths / pop_estimate * 100000
  ) %>%
  mutate(
    cases_per100k = cases / pop_estimate * 100000
  )

## 5) arrange in descending order by cases_per100k
## then take the top 10 
## 5b) I noticed the results are different by deaths than by cases
## so I will do both results, unlike the example table which arranges by cases

top10_cases_per100k <- top10_per100k %>%
  arrange(
    desc(cases_per100k)
  ) %>%
  head(n = 10)
as_tibble(top10_cases_per100k)

top10_deaths_per100k <- top10_per100k %>%
  arrange(
    desc(deaths_per100k)
  ) %>%
  head(n = 10)
as_tibble(top10_deaths_per100k)

## 6) print out the results 
writeLines("10 states with highest cases per 100,000 people:")
print(top10_cases_per100k$state)
writeLines("10 cases with highest number of deaths per 100,000 people:")
print(top10_deaths_per100k$state)

```
-- Communicate your methodology, results, and interpretation here -- 
These results are very different from the results of the first data set which 
only calculated the total cases and deaths. This data set is much more valuable 
for researching a virus because it tells us the rate of infection and tells 
us where and how we need to allocate resources as well as how the population is 
responding to measures like vaccines and social distancing for COVID 
specifically. Additionally the other table may under estimate the rate of 
infection of smaller states because the total is simply lower due to population.
I think this statistical approach that uses per capita is much more appropriate
because a raw total just reiterates what we already know that there are more 
people in states like California, Texas, and New York. 

##### Question 3

Now, select a state and calculate the seven-day averages for new cases and 
deaths per 100,000 people. Once you have calculated the averages, create a
visualization using ggplot2 to represent the data. 

```{r, part2_question3, echo=TRUE, include=TRUE}
## 1) of course I have to pick California my home state
## 2) first calculate the number of new cases 
california_only <- counties_combined %>%
  filter(state == "California") 

california_by_new <- california_only %>%
  group_by(date) %>%
  summarise(
    deaths_CA = sum(deaths),
    cases_CA = sum(cases)
  )
## 3) ^^ this gave us the total but now let's calculate the difference 
## between each date using a for loop
## 3b) first set up an empty vector of double class to store the values
## 3c) calculate the difference using a for loop 
## 3d) add the resulting vector to our data set 
  
new_CA_deaths <- vector("double", length(california_by_new$deaths_CA))

for(i in 1:length(california_by_new$deaths_CA)){
 if(i <= 1){ 
    new_CA_deaths <- (california_by_new$deaths_CA[i])
 }
  else{
    new_CA_deaths <- append(new_CA_deaths,
            california_by_new$deaths_CA[i] - california_by_new$deaths_CA[i - 1])
  }
}

california_by_new$new_CA_deaths <- new_CA_deaths

## 4) Repeat for cases

new_CA_cases <- vector("double", length(california_by_new$cases_CA))
for(i in 1:length(california_by_new$cases_CA)){
 if(i <= 1){ 
    new_CA_cases <- (california_by_new$cases_CA[i])
 }
  else{
    new_CA_cases <- append(new_CA_cases,
            california_by_new$cases_CA[i] - california_by_new$cases_CA[i - 1])
  }
}

california_by_new$new_CA_cases <- new_CA_cases

## 5) now calculate the 7-day average using rollmeanr
## specify for NA values to be filled
california_by7 <- 
  transform(california_by_new,
    avg7_deaths = rollmeanr(new_CA_deaths, 7, fill = NA)
  )

california_by7 <- 
  transform(california_by7, 
            avg7_cases = rollmeanr(new_CA_cases, 7, fill = NA))

## 6) calculate the 7-day avg for each variable per 100,000 people 
## utilize the population from before but filter for only the CA values 
## 6b) set up the last dates for each year to define the limits we'll use
CA_pop_est <- state_pop_est %>%
  filter(state == "California")

last_day2020 <- as.Date('2020-12-31')
last_day2021 <- as.Date('2021-12-31')

## 6c) find the last day of 2020 in california_by7

last_day2020 <- which(california_by7 == "2020-12-31", arr.ind = TRUE)

## it is in row 342 column 1, we will use "last_day[1, 1]" through i 

## 7) use the pop. est in a for loop to check the date, we can 
## use the pop. est of 2021 to calculate values for 2022, but just know 
## that the data may be over representing the number of cases and deaths 
## per 100k people as we can assume the pop. of CA would be higher in 2022

#####################################################
#####################################################
#####################################################

cases_CA_7day_100k <- vector("double", length(california_by7$avg7_cases))
avg7_cases <- california_by7$avg7_cases
for( i in 1: length(california_by7$avg7_cases)){
  if( i <= last_day2020[1,1]){
    cases_CA_7day_100k[i] <- (avg7_cases[i] / (CA_pop_est[1,3]) * 100000)[[1]][1]
  }
  else{
    cases_CA_7day_100k[i] <- (avg7_cases[i] / (CA_pop_est[2,3]) * 100000)[[1]][1]
  }
}
## print(cases_CA_7day_100k)


## 7b) repeat for deaths
deaths_CA_7day_100k <- vector("double", length(california_by7$avg7_deaths))
avg7_deaths <- california_by7$avg7_deaths

for(i in 1: length(california_by7$avg7_deaths)){
  if(i <= last_day2020[1,1]){
    deaths_CA_7day_100k[i] <- (avg7_deaths[i] / (CA_pop_est[1,3]) * 100000)[[1]][1]
  }
  else{
    deaths_CA_7day_100k[i] <- (avg7_deaths[i] / CA_pop_est[2,3] * 100000)[[1]][1]
  }
}
## print(deaths_CA_7day_100k)

## 8) add vectors to data set 
CA_7day_100k <- california_by7
CA_7day_100k$cases_7day_100k <- cases_CA_7day_100k
CA_7day_100k$deaths_7day_100k <- deaths_CA_7day_100k

## 8) create a plot data set with only the two variables we need
## change to two columns to use facet_wrap easily 
CA_plot <- CA_7day_100k %>%
  select(date, cases_7day_100k, deaths_7day_100k) %>%
  pivot_longer(c('cases_7day_100k','deaths_7day_100k'),
               names_to = 'type',
               values_to = 'number_reported'
  )

## 8b) 
variable_names <- list(
  "cases_7day_100k" = "Cases (per 100,000 people)" ,
  "deaths_7day_100k" = "Deaths (per 100,000 people)"
)

variable_labeller <- function(variable, value){
  return(variable_names[value])
}

## 9) now create a visualization
CA_plot_2 <- CA_plot %>% ggplot(
  aes(x = date,
      y = number_reported,
      fill = type)) +
         geom_bar(stat = 'identity',
                  size = .25, na.rm = TRUE)+
    facet_wrap(~ type,
             nrow =  2, 
             scales = "free",
             labeller = variable_labeller) +
  labs(x = "Date", 
       y = "7-day avg. Deaths       7-day avg. Cases", 
       fill = "Type Reported:") +
  theme(title = element_text(face = "bold",
                             size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_text(face = "bold")) +
  ggplot2::labs(title =
              "        California COVID-19 Deaths & Cases:
              7-day avg. per 100,000 people") + 
  scale_fill_discrete(labels = c("Cases", "Deaths")) +
  scale_x_date(date_labels = "%b %Y", 
               date_breaks = "3 months",
               limits = as.Date(c('2020-03-15','2022-10-15')))
CA_plot_2

```
-- Communicate your methodology, results, and interpretation here -- 
First, we found the number of new cases & deaths per day for California
via a for loop method for each variable. I did this 
without setting date limits just to see the results for all of COVID-19 
history in California. Then, we used that new data set to find the 7-day rolling
average of new cases & deaths in california using the zoo package rollmeanr(). 
Next we used another for loop to calculate the avg. per 100k people, which was
odd because my for loop was storing my results in a listed column, therefore
nesting the results, but then I figured out if I just called the value 
for each calculation then store it into the vector that it worked correctly. 
Then finally we used the data set variables that we needed to create a 
plot data set, set up to make using facet_wrap faster. Finally we plotted
and we can see the that highest number of deaths and cases do have a strong 
correlation across time. We can also see that the number of cases and deaths 
in California strongly resembles our first plot for the whole United States, 
except for the first initial spike in April 2020 for deaths that we saw 
on the plot for the whole US. I used the same population from 2021 for CA
as for 2022 because the data set only went up to 2021, so it may be over
representing the number of cases and deaths. I see there are values 
that are negative in the deaths due to the difference in cases being less
between the two successive dates, which could be misleading and confusing, 
but perhaps we could account for this by setting an extra condition 
while calculating the new deaths and cases, where is there is a negative
difference to just set it to zero, but that would be changing the data we
were provided. I double checked the total number of cases and deaths for 
COVID-19 in CA all-time and the numbers are correct so I would assume or
data point for each date are also correct. 


##### Question 4

Using the same state, identify the top 5 counties in terms of deaths and 
cases per 100,000 people. 
```{r, part2_question4, echo=TRUE, include=TRUE}
## 1) filter for the last date of 2021 because that is when we 
## have the pop. est.
max_date = as.Date('2021-12-31')

CA_counties <- california_only %>%
  filter(date == as.Date(max_date)) %>%
  arrange(desc(deaths), cases)

## 2) get the pop. est. 
CA_counties_pop <- us_population_estimates %>%
  filter(STNAME == "California") %>%
  filter(Year == 2021) %>%
  rename(county = CTYNAME) %>%
  select(county, Year, Estimate)

## 2b) the counties all have the word county after it which needs to be changed
## use gsub() to select the pattern and replace it with nothing in the column

CA_counties_pop$county <- 
  gsub("County", "", as.character(CA_counties_pop$county))


## 3) apply the pop_est to the data set to calculate the cases and deaths 
## per 100,000 people, just add the columns in to the data set, make sure 
## they are in alphabetical order to align them correctly
CA_counties <- CA_counties %>% 
  arrange(county) %>%
  select(state, county, cases, deaths, fips)

CA_counties$year <- CA_counties_pop$Year
CA_counties$estimate <- CA_counties_pop$Estimate

## 4) calculate the number of cases and deaths per 100k 

CA_counties <- CA_counties %>%
  mutate(
    cases_per100k = cases / estimate * 100000
  ) %>%
  mutate(
    deaths_per100k = deaths / estimate * 100000
  )

CA_counties_5 <- CA_counties %>%
  arrange(desc(cases_per100k), deaths_per100k) %>%
  head(n = 5)

as_tibble(CA_counties_5)

writeLines("Top 5 counties in California with highest total 
           cases and deaths per 100,000 people as of Dec. 31 2021:
           (in desceding order)")
print(CA_counties_5$county)
  
```

-- Communicate your methodology, results, and interpretation here -- 
Well, this was interesting for me to see as a California native. The
top 5 counties per 100k is strikingly different than the total cases and 
deaths ranking. The only county with a large population among the top 5 
is San Bernardino, but the big counties like LA, San Diego, Orange, and 
Santa Clara aren't aren't on there and even LA isn't on the list until #7. 
I decided to use Dec 31 2021 as the date for analysis because it would have
the most accurate results since we only have the pop. estimates up to 2021 
AND we used total cases and deaths not the new cases and deaths. I wonder
if the results would be different if calculates for 2020 and 2022. 


##### Question 5

Modify the code below for the map projection to plot county-level deaths and 
cases per 100,000 people for your state. 

```{r, part2_question5, echo=TRUE, include=TRUE}

plot_CA <- plot_usmap(regions = "county",
           include = "California",
           data = CA_counties,
           values = "deaths_per100k", 
           coor = 'blue', 
           labels = FALSE,
           label_color = "black",
           color = "black",
           linewidth = 2)+
    scale_fill_continuous(low = 'white', 
                        high = 'blue',
                        name = "Deaths per 100,000")+
  theme(legend.position = c(.99,.4),
        legend.background = element_rect(fill = "grey",
                                         size = .5, 
                                         linetype = "solid",
                                         color = "darkblue"),
        legend.text = element_text(color = "black", face = "bold"),
        title = element_text(face = "bold"))+
  labs(title = "California Counties COVID-19 Deaths per 100k People:",
        subtitle = "       based on 2021 reports and population estimates")

plot_CA
  
  
```
-- Communicate your methodology, results, and interpretation here -- 
Alright, this is pretty straight forward we just plugged in the 
data set we made in the previous question. I tried including the labels but 
it made the map overwhelming. I am not surprised that Imperial County has the 
highest deaths per 100k because it is a very poor county where the poverty rate 
is about 25%. I think that in counties like that there were less COVID resources
like vaccines. An interesting test would be to see if there is a correlation 
between poverty rates and COVID deaths by county. 


##### Question 6

Finally, select three other states and calculate the seven-day averages 
for new deaths and cases per 100,000 people for between March 15, 2020, 
and December 31, 2021. 

```{r, part2_question6, echo=TRUE, include=TRUE}
## 1) I will pick Colorado, Maryland, and New York
###############################################################
###############################################################
## 2) COLORADO 
## find the difference in deaths and cases
CO_counties_diff <- counties_combined %>%
  filter(state == "Colorado") %>%
  filter(date >= as.Date("2020-03-15")) %>%
  filter(date <= as.Date("2021-12-31")) %>%
  group_by(date) %>%
  summarise(
    deaths_CO = sum(deaths),
    cases_C0 = sum(cases)
  )

new_CO_deaths <- vector("double", length(CO_counties_diff$deaths_CO))

for(i in 1:length(CO_counties_diff$deaths_CO)){
 if(i <= 1){ 
    new_CO_deaths <- (CO_counties_diff$deaths_CO[i])
 }
  else{
    new_CO_deaths <- append(new_CO_deaths,
            CO_counties_diff$deaths_CO[i] - CO_counties_diff$deaths_CO[i - 1])
  }
}

CO_counties_diff$new_deaths_CO<- new_CO_deaths

new_CO_cases <- vector("double", length(CO_counties_diff$cases_C0))
for(i in 1:length(CO_counties_diff$cases_C0)){
 if(i <= 1){ 
    new_CO_cases <- (CO_counties_diff$cases_C0[i])
 }
  else{
    new_CO_cases <- append(new_CO_cases,
            CO_counties_diff$cases_C0[i] - CO_counties_diff$cases_C0[i - 1])
  }
}

CO_counties_diff$new_cases_C0 <- new_CO_cases

## use the new cases and deaths to find the 7day avg. 

CO_by7 <- 
  transform(CO_counties_diff,
    avg7_deaths = rollmeanr(new_cases_C0, 7, fill = NA)
  )

CO_by7 <- 
  transform(CO_by7, 
            avg7_cases = rollmeanr(new_deaths_CO, 7, fill = NA))

## use the 7-day avg. to calc. the value per 100k people
## which will depend on the year 
CO_pop_year <- state_pop_est %>%
  filter(state == "Colorado") 

cases_CO_7day_100k <- vector("double", length(CO_by7$avg7_cases))
avg7_cases <- CO_by7$avg7_cases
for( i in 1: length(CO_by7$avg7_cases)){
  if( i <= last_day2020[1,1]){
    cases_CO_7day_100k[i] <- (avg7_cases[i] / (CO_pop_year[1,3]) * 100000)[[1]][1]
  }
  else{
    cases_CO_7day_100k[i] <- (avg7_cases[i] / (CO_pop_year[2,3]) * 100000)[[1]][1]
  }
}
## print(cases_CO_7day_100k)
## 7b) repeat for deaths
deaths_CO_7day_100k <- vector("double", length(CO_by7$avg7_deaths))
avg7_deaths <-CO_by7$avg7_deaths

for(i in 1: length(CO_by7$avg7_deaths)){
  if(i <= last_day2020[1,1]){
    deaths_CO_7day_100k[i] <- (avg7_deaths[i] / (CO_pop_year[1,3]) * 100000)[[1]][1]
  }
  else{
    deaths_CO_7day_100k[i] <- (avg7_deaths[i] / CO_pop_year[2,3] * 100000)[[1]][1]
  }
}
## print(deaths_C0_7day_100k)
## add to data set
CO_7day_100k <- CO_by7
CO_7day_100k$cases_7day_100k <- cases_CO_7day_100k
CO_7day_100k$deaths_7day_100k <- deaths_CO_7day_100k
## print main results
as_tibble(CO_7day_100k)%>%
  select(date, cases_7day_100k, deaths_7day_100k)

###############################################################
###############################################################
## MARYLAND

MD_counties_diff <- counties_combined %>%
  filter(state == "Maryland") %>%
  filter(date >= as.Date("2020-03-15")) %>%
  filter(date <= as.Date("2021-12-31")) %>%
  group_by(date) %>%
  summarise(
    deaths_MD = sum(deaths),
    cases_MD = sum(cases)
  )

new_MD_deaths <- vector("double", length(MD_counties_diff$deaths_MD))

for(i in 1:length(MD_counties_diff$deaths_MD)){
 if(i <= 1){ 
    new_MD_deaths <- (MD_counties_diff$deaths_MD[i])
 }
  else{
    new_MD_deaths <- append(new_MD_deaths,
            MD_counties_diff$deaths_MD[i] - MD_counties_diff$deaths_MD[i - 1])
  }
}

MD_counties_diff$new_deaths_MD<- new_MD_deaths

new_MD_cases <- vector("double", length(MD_counties_diff$cases_MD))
for(i in 1:length(MD_counties_diff$cases_MD)){
 if(i <= 1){ 
    new_MD_cases <- (MD_counties_diff$cases_MD[i])
 }
  else{
    new_MD_cases <- append(new_MD_cases,
            MD_counties_diff$cases_MD[i] - MD_counties_diff$cases_MD[i - 1])
  }
}

MD_counties_diff$new_cases_MD <- new_MD_cases
## use the new cases and deaths to find the 7day avg. 

MD_by7 <- 
  transform(MD_counties_diff,
    avg7_deaths = rollmeanr(new_cases_MD, 7, fill = NA)
  )

MD_by7 <- 
  transform(MD_by7, 
            avg7_cases = rollmeanr(new_deaths_MD, 7, fill = NA))

## use the 7-day avg. to calc. the value per 100k people
## which will depend on the year 
MD_pop_year <- state_pop_est %>%
  filter(state == "Maryland") 

cases_MD_7day_100k <- vector("double", length(MD_by7$avg7_cases))
avg7_cases <- MD_by7$avg7_cases
for( i in 1: length(MD_by7$avg7_cases)){
  if( i <= last_day2020[1,1]){
    cases_MD_7day_100k[i] <- (avg7_cases[i] / (MD_pop_year[1,3]) * 100000)[[1]][1]
  }
  else{
    cases_MD_7day_100k[i] <- (avg7_cases[i] / (MD_pop_year[2,3]) * 100000)[[1]][1]
  }
}
## print(cases_MD_7day_100k)
## 7b) repeat for deaths
deaths_MD_7day_100k <- vector("double", length(MD_by7$avg7_deaths))
avg7_deaths <-MD_by7$avg7_deaths

for(i in 1: length(MD_by7$avg7_deaths)){
  if(i <= last_day2020[1,1]){
    deaths_MD_7day_100k[i] <- (avg7_deaths[i] / (MD_pop_year[1,3]) * 100000)[[1]][1]
  }
  else{
    deaths_MD_7day_100k[i] <- (avg7_deaths[i] / MD_pop_year[2,3] * 100000)[[1]][1]
  }
}
## print(deaths_MD_7day_100k)
## add to data set
MD_7day_100k <- MD_by7
MD_7day_100k$cases_7day_100k <- cases_MD_7day_100k
MD_7day_100k$deaths_7day_100k <- deaths_MD_7day_100k
## print main results
as_tibble(MD_7day_100k) %>%
  select(date, cases_7day_100k, deaths_7day_100k)

###############################################################
###############################################################
## NEW YORK
NY_counties_diff <- counties_combined %>%
  filter(state == "New York") %>%
  filter(date >= as.Date("2020-03-15")) %>%
  filter(date <= as.Date("2021-12-31")) %>%
  group_by(date) %>%
  summarise(
    deaths_NY = sum(deaths),
    cases_NY = sum(cases)
  )

new_NY_deaths <- vector("double", length(NY_counties_diff$deaths_NY))

for(i in 1:length(NY_counties_diff$deaths_NY)){
 if(i <= 1){ 
    new_NY_deaths <- (NY_counties_diff$deaths_NY[i])
 }
  else{
    new_NY_deaths <- append(new_NY_deaths,
            NY_counties_diff$deaths_NY[i] - NY_counties_diff$deaths_NY[i - 1])
  }
}

NY_counties_diff$new_deaths_NY<- new_NY_deaths

new_NY_cases <- vector("double", length(NY_counties_diff$cases_NY))
for(i in 1:length(NY_counties_diff$cases_NY)){
 if(i <= 1){ 
    new_NY_cases <- (NY_counties_diff$cases_NY[i])
 }
  else{
    new_NY_cases <- append(new_NY_cases,
            NY_counties_diff$cases_NY[i] - NY_counties_diff$cases_NY[i - 1])
  }
}

NY_counties_diff$new_cases_NY <- new_NY_cases
## use the new cases and deaths to find the 7day avg. 

NY_by7 <- 
  transform(NY_counties_diff,
    avg7_deaths = rollmeanr(new_cases_NY, 7, fill = NA)
  )

NY_by7 <- 
  transform(NY_by7, 
            avg7_cases = rollmeanr(new_deaths_NY, 7, fill = NA))
## use the 7-day avg. to calc. the value per 100k people
## which will depend on the year 
NY_pop_year <- state_pop_est %>%
  filter(state == "New York") 

cases_NY_7day_100k <- vector("double", length(NY_by7$avg7_cases))
avg7_cases <- NY_by7$avg7_cases
for( i in 1: length(NY_by7$avg7_cases)){
  if( i <= last_day2020[1,1]){
    cases_NY_7day_100k[i] <- (avg7_cases[i] / (NY_pop_year[1,3]) * 100000)[[1]][1]
  }
  else{
    cases_NY_7day_100k[i] <- (avg7_cases[i] / (NY_pop_year[2,3]) * 100000)[[1]][1]
  }
}
## print(cases_NY_7day_100k)
## 7b) repeat for deaths
deaths_NY_7day_100k <- vector("double", length(NY_by7$avg7_deaths))
avg7_deaths <-NY_by7$avg7_deaths

for(i in 1: length(NY_by7$avg7_deaths)){
  if(i <= last_day2020[1,1]){
    deaths_NY_7day_100k[i] <- (avg7_deaths[i] / (NY_pop_year[1,3]) * 100000)[[1]][1]
  }
  else{
    deaths_NY_7day_100k[i] <- (avg7_deaths[i] / NY_pop_year[2,3] * 100000)[[1]][1]
  }
}
## print(deaths_CA_7day_100k)
## add to data set
NY_7day_100k <- NY_by7
NY_7day_100k$cases_7day_100k <- cases_NY_7day_100k
NY_7day_100k$deaths_7day_100k <- deaths_NY_7day_100k
## print main results
as_tibble(NY_7day_100k)%>%
  select(date, cases_7day_100k, deaths_7day_100k)

```
-- Communicate your methodology, results, and interpretation here -- 
We really just used the same methods for the calculations for California


##### Question 7
Create a visualization comparing the seven-day averages for new deaths and 
cases per 100,000 people for the four states you selected. 

```{r, part2_question7, echo = TRUE, include=TRUE}
## Create a visualization comparing the seven-day averages for new deaths and 
## cases per 100,000 people for the four states you selected. 

## CO #########################################
## create a plot data set with only the two variables we need
## change to two columns to use facet_wrap easily 
CO_plot <- CO_7day_100k %>%
  select(date, cases_7day_100k, deaths_7day_100k) %>%
  pivot_longer(c('cases_7day_100k','deaths_7day_100k'),
               names_to = 'type',
               values_to = 'number_reported'
  )


variable_names <- list(
  "cases_7day_100k" = "Cases (per 100,000 people)" ,
  "deaths_7day_100k" = "Deaths (per 100,000 people)"
)

variable_labeller <- function(variable, value){
  return(variable_names[value])
}

CO_plot2 <- CO_plot %>% ggplot(
  aes(x = date,
      y = number_reported,
      fill = type)) +
         geom_bar(stat = 'identity',
                  size = .25, na.rm = TRUE)+
    facet_wrap(~ type,
             nrow =  2, 
             scales = "free",
             labeller = variable_labeller) +
  labs(x = "Date", 
       y = "7-day avg. Deaths       7-day avg. Cases", 
       fill = "Type Reported:") +
  theme(title = element_text(face = "bold",
                             size = 7),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_text(face = "bold")) +
  ggplot2::labs(title =
              "        Colorado COVID-19 Deaths & Cases:
              7-day avg. per 100,000 people") + 
  scale_fill_discrete(labels = c("Cases", "Deaths")) +
  scale_x_date(date_labels = "%b %Y", 
               date_breaks = "3 months",
               limits = as.Date(c('2020-03-15','2021-12-31')))

## MD #########################################
MD_plot <- MD_7day_100k %>%
  select(date, cases_7day_100k, deaths_7day_100k) %>%
  pivot_longer(c('cases_7day_100k','deaths_7day_100k'),
               names_to = 'type',
               values_to = 'number_reported'
  )
MD_plot2 <- MD_plot %>% ggplot(
  aes(x = date,
      y = number_reported,
      fill = type)) +
         geom_bar(stat = 'identity',
                  size = .25, na.rm = TRUE)+
    facet_wrap(~ type,
             nrow =  2, 
             scales = "free",
             labeller = variable_labeller) +
  labs(x = "Date", 
       y = "7-day avg. Deaths       7-day avg. Cases", 
       fill = "Type Reported:") +
  theme(title = element_text(face = "bold", 
                             size = 6),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_text(face = "bold")) +
  ggplot2::labs(title =
              "        Maryland COVID-19 Deaths & Cases:
              7-day avg. per 100,000 people") + 
  scale_fill_discrete(labels = c("Cases", "Deaths")) +
  scale_x_date(date_labels = "%b %Y", 
               date_breaks = "3 months",
               limits = as.Date(c('2020-03-15','2021-12-31')))


## NY
NY_plot <- NY_7day_100k %>%
  select(date, cases_7day_100k, deaths_7day_100k) %>%
  pivot_longer(c('cases_7day_100k','deaths_7day_100k'),
               names_to = 'type',
               values_to = 'number_reported')

NY_plot2 <- NY_plot %>% ggplot(
  aes(x = date,
      y = number_reported,
      fill = type)) +
         geom_bar(stat = 'identity',
                  size = .25, na.rm = TRUE)+
    facet_wrap(~ type,
             nrow =  2, 
             scales = "free",
             labeller = variable_labeller) +
  labs(x = "Date", 
       y = "7-day avg. Deaths       7-day avg. Cases", 
       fill = "Type Reported:") +
  theme(title = element_text(face = "bold",
                             size = 6),
        axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_text(face = "bold")) +
  ggplot2::labs(title =
              "        New York COVID-19 Deaths & Cases:
              7-day avg. per 100,000 people") + 
  scale_fill_discrete(labels = c("Cases", "Deaths")) +
  scale_x_date(date_labels = "%b %Y", 
               date_breaks = "3 months",
               limits = as.Date(c('2020-03-15','2021-12-31')))

plot_grid(CA_plot_2, CO_plot2)
plot_grid(MD_plot2, NY_plot2)

```

-- Communicate your methodology, results, and interpretation here -- 
We just repeated the same process for California. One thing 
I did notice about the NY data set is that the COVID data set does not 
include the Boroughs of New York like Kings, Bronx, etc. so this data
may be under representing the total COVID cases and deaths because 
even when I summed up the population for NY based on the counties present in 
the data set the total population came up significantly lower than the 
expected 19 million. The facet_wrap, grid_arrange(), and even plot_grid 
could not handle all 4 plots at once. When I tried to use them to compare
side by side the value and meanings were lost. Overall each state does 
seem to show a strong two spike correlation around the COVID variants times, but 
California's is closest to the US total, the Colorado results per 100k 
shows a higher rate than for deaths and cases overall than the other states. 
Maryland has the most unusual results that don't follow as much of a 
pattern that we would expect, the cases spike earlier than expected for 
both spikes. Then New York overall shows much lower rates for both cases and 
deaths per 100k people. I think if we had all the data including the Boroughs
the results would be different because there was likely more COVID reports
and there populations of each is very high. 




